# Constant and Variable 
Constant are variable, include numerical constant or String constant etc. 

Numerical cosntant can be integer or decimal point number. R will automatically switch to proper data type.

```{r constant_variable, results='hide'}
123
3.14
-0.02
1.23E2
1.2E-2
123L # the L suffix indicates an integer
```



Variable is used to store data, and can be created by assignment operator `<-` or `=`.

Variable name can be composed of letters, numbers, dot (.) and underscore (_), but cannot start with number or underscore. Variable names are case-sensitive.

```{r variable_example, results='hide'}
a <- 10
b = 3.14
my_var <- "Hello, R!"

tax.table <- data.frame(ID = 1:5, Name = c("A", "B", "C", "D", "E")) # not recommended to use. # nolint
tax.name <- "Tax Data"
```


# Numerical vector and calculation
use `c()` function to create a vector, which is a one-dimensional array that can hold multiple values of the same type.


```{r vector_creation}

marks <- c(85, 90, 78, 92, 88)
x1 <- c(1:3, 10:13)
x2 <- c(20:25)
x <- c(x1, x2) # 这个是拼接，不是二维向量
print(x)
```

If two vector have the same length, the calculation between them is element-wise:
```{r vector_calculation}
x1 <- c(1, 10)
x2 <- c(4, 2)
x1 + x2
## [1]  5 12
x1 - x2
## [1] -3  8
x1 * x2
## [1]  4 20
x1 / x2
## [1] 0.25 5.00
```

If two vector have different length, the shorter one will be recycled to match the length of the longer one:
```{r vector_recycling}
a <- c(1, 2, 3, 4)
b <- c(4, 5)
a + b
## [1] 5 7 7 9
```

When function is applied to a vector, it will be applied to each element of the vector:

```{r vector_function}
nums <- c(1, 4, 9, 16)
sqrt_nums <- sqrt(nums)
print(sqrt_nums) # prints c(1, 2, 3, 4
```


some common mathematical functions that can be applied to vectors, can run code `help.start()` in R console to open the help page and search for more functions.

Commonly used mathematical functions:

| Category                          | Functions                                                  |
| --------------------------------- | ---------------------------------------------------------- |
| Rounding                          | `ceiling`, `floor`, `round`, `signif`, `trunc`, `zapsmall` |
| Sign function                     | `sign`                                                     |
| Absolute value                    | `abs`                                                      |
| Square root                       | `sqrt`                                                     |
| Logarithm & exponential functions | `log`, `exp`, `log10`, `log2`                              |
| Trigonometric functions           | `sin`, `cos`, `tan`                                        |
| Inverse trigonometric functions   | `asin`, `acos`, `atan`, `atan2`                            |
| Hyperbolic functions              | `sinh`, `cosh`, `tanh`                                     |
| Inverse hyperbolic functions      | `asinh`, `acosh`, `atanh`                                  |

Less commonly used mathematical functions:

| Category                         | Functions                                                            |
| -------------------------------- | -------------------------------------------------------------------- |
| Beta functions                   | `beta`, `lbeta`                                                      |
| Gamma functions                  | `gamma`, `lgamma`, `digamma`, `trigamma`, `tetragamma`, `pentagamma` |
| Combinatorics                    | `choose`, `lchoose`                                                  |
| Fourier transforms & convolution | `fft`, `mvfft`, `convolve`                                           |
| Orthogonal polynomials           | `poly`                                                               |
| Root finding                     | `polyroot`, `uniroot`                                                |
| Optimization                     | `optimize`, `optim`                                                  |
| Bessel functions                 | `besselI`, `besselK`, `besselJ`, `besselY`                           |
| Spline interpolation             | `spline`, `splinefun`                                                |
| Simple differentiation           | `deriv`                                                              |


# Sort and order
You can use `sort()` function to sort a vector in ascending or descending order. The `rev(x)` function can be used to reverse the order of a vector. `order(x)` function return the indices that would sort the vector.

```{r sort_order_example}
vec <- c(5, 2, 8, 1, 4)
sorted_vec <- sort(vec) # ascending order
cat("the vector sorted in ascending order:", sorted_vec, "\n")
desc_sorted_vec <- sort(vec, decreasing = TRUE) # descending order
cat("the vector sorted in descending order:", desc_sorted_vec, "\n")
reversed_vec <- rev(vec) # reverse order
cat("the vector in reverse order:", reversed_vec, "\n")
order_indices <- order(vec) # indices that would sort the vector
cat("the indices that would sort the vector:", order_indices, "\n")
sort_by_indices <- vec[order_indices] # sort using the indices
cat("the vector sorted using the order indices:", sort_by_indices, "\n")
```

# Statistical functions

commonly used statistical functions that can be applied to vectors: like `mean()`, `median()`, `sd()`, `var()`, `min()`, `max()`, `quantile()`, `summary()` etc.

```{r statistical_functions_example}
var <- c(10, 20, 30, 40, 50, rnorm(3))
print(var)
print(mean(var))      # calculates the mean
print(median(var))    # calculates the median
print(sd(var))        # calculates the standard deviation
print(var(var))       # calculates the variance
print(min(var))       # finds the minimum value
print(max(var))       # finds the maximum value
print(quantile(var, probs = seq(0, 1, 0.25)))  # calculates the quantiles
print(summary(var))   # provides a summary of statistics
```